Interrogating the linear variable importance of local explanations of black-box models with animated linear projections

### 2021-12-03 _paper.rmd
images to .png; to allow for printing .pdf (.pdf format with alpha causes `flattening` to hang).

<!-- ## XAI & interpretability crisis -->
Black-box models are becoming increasingly common, but not without their share of controversy and issues [@oneil_weapons_2016, @kodiyan_overview_2019]. Applications have been known to reflect common biases against sex [@dastin_amazon_2018, @duffy_apple_2019], race [@larson_how_2016], and age [@diaz_addressing_2018]. These are all-too common issue, when endogenous or exogenous biases in the training data and identified and mimicked by models. Another issue is that of data-drift when new data is outside the support of latent or exogenous explanatory variables. Data drift can lead to worse predictions [@lazer_parable_2014, @salzberg_why_2014]. Such issues highlight the need to make models fair, accountable, ethical, and transparent which has led to the movement of XAI [@adadi_peeking_2018, @arrieta_explainable_2020].

@@

<!-- ## manual tours --> 
There are various types of tours, which are distinguished by the sequence of projection bases. In a _manual_ tour [@cook_manual_1997, @spyrison_spinifex_2020] this path is defined by changing the contribution of a selected variable. <!--tours and models --> Applying tours in conjunction with models has been previously done,  _ie_ for exploring various statistical model fits [@wickham_visualizing_2015], and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis path [@lee_pptree_2013, @da_silva_projection_2021].

@@

```{r shapdistrbd, echo=F, fig.cap = "Illustration of the distribution of SHAP attributions, the SHAP values, and a breakdown plot, the typical visual of SHAP local explanations. For FIFA 2020 data, of a random forest model regressing wages from 9 skill attributes for a star offensive and defensive player. a) The players have very different wages. b) Shows the distributions of the attributions permuting over 25 permutations in the explanatory variables. The median of these distributions are the final SHAP values, notice that that the variable importance differs across the exogenous information of player position. These explanations make sense; the variable importances make sense in light of the position of the player. c) Breakdown plots of the observations the explanation used to additively explain the difference between the intercept and prediction"}
knitr::include_graphics("./figures/shap_distr_bd.pdf")
```

@@

The regression case is not as discrete a feature. Instead, we find it more meaningful to compare the observed value or  residual with the projection. In this case the horizontal positions

@@

<!-- Dipiction of a frame -->
The SHAP values of the full dataset are shown as horizontal parrallel coordiate plot at the top with the selected observations highlighted. In the classification case, 1D density curves with underlying rug marks are drawn and colored according to their predicted classes. In the regression case, the horizontal position of the points comes from projection through the 1D attribution basis while the vertical position of is fixed to the observations prediction and residual respectively. The current projection basis is depicted on the bottom, where the width of the bar is mapped to variables contribution to the horizontal axis. The starting frame is the normalized SHAP values of the primary observation. In figure \@ref(fig:cheemtourclass) we illustrate the classification case with simulated data; 3 clusters separated in the first 2 dimensions with 2 dimensions of noise.

@@

```{r casepenguins, echo=F, out.width = "100%", fig.cap = "Species classification of Palmer penguin data."}
knitr::include_graphics("./figures/case_penguins.pdf")
```

@@

```{r casechocolates, echo=F, out.width = "100%", fig.cap = "Chocolates data  type classification (milk or dark). "}
knitr::include_graphics("./figures/case_chocolates.pdf")
```

@@

```{r casefifa, echo=F, out.width = "100%", fig.cap = "FIFA 2020, regressing log wages [2020 Euros] from aggregations of skill measurements. The primary observation is star offensive player (L. Messi) as compared with a top defensive player (V. van Dijk)."}
knitr::include_graphics("./figures/case_fifa.pdf")
```

@@

```{r caseames, echo=F, out.width = "100%", fig.cap = "Ames housing 2018 regressing log sales price [2018 USD]."}
knitr::include_graphics("./figures/case_ames2018.pdf")
```

>>>>

<!-- ## XAI & interpretability crisis -->
Black-box models are becoming increasingly common, but not without their share of controversy and issues [@oneil_weapons_2016, @kodiyan_overview_2019]. Applications have been known to reflect common biases against sex [@dastin_amazon_2018, @duffy_apple_2019], race [@larson_how_2016], and age [@diaz_addressing_2018]. These are all-too common of an issue, when endogenous or exogenous biases in the training data and identified and mimicked by models. Another issue is that of data drift when new data is outside the support of latent or exogenous explanatory variables. Data drift can lead to worse predictions [@lazer_parable_2014, @salzberg_why_2014]. Such issues highlight the need to make models fair, accountable, ethical, and transparent which has led to the movement of XAI [@adadi_peeking_2018, @arrieta_explainable_2020].

@@

<!-- ## manual tours --> 
There are various types of tours, which are distinguished by the sequence of projection bases. In a _manual_ tour [@cook_manual_1997, @spyrison_spinifex_2020] this path is defined by changing the contribution of a selected variable. <!--tours and models --> Applying tours in conjunction with models has been done in a couple contexts, _ie_ for exploring various statistical model fits [@wickham_visualizing_2015], and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis path [@lee_pptree_2013, @da_silva_projection_2021].

@@

```{r shapdistrbd, echo=F, fig.cap = "Illustration of the distribution of SHAP attributions, the SHAP values, and a breakdown plot, the typical visual of SHAP local explanations. For FIFA 2020 data, of a random forest model regressing wages from 9 skill attributes for a star offensive and defensive player. a) The players have very different wages. b) Shows the distributions of the attributions permuting over 25 permutations in the explanatory variables. The median of these distributions are the final SHAP values, notice that that the variable importance differs across the exogenous information of player position. These explanations make sense; the variable importances make sense in light of the position of the player. c) Breakdown plots of the observations the explanation used to additively explain the difference between the intercept and prediction"}
knitr::include_graphics("./figures/shap_distr_bd.png")
```

@@

The regression case is not as discrete a feature. Instead, we find it more meaningful to compare the observed value or  residual with the projection. In this case the horizontal positions are the same, that is projection of the data through the attribution of the explanation and then varying as the radial tour changes the contributions of the selected variable. While the height of the points is fixed to the fixed either to the observed y and residuals for the middle and right panels respectively.

@@

<!-- Dupiction of a frame -->
The SHAP values of the full dataset are shown as horizontal parrallel coordiate plot at the top with the selected observations highlighted. In the classification case, 1D density curves with underlying rug marks are drawn and colored according to their predicted classes. In the regression case, the horizontal position of the points comes from projection through the 1D attribution basis while the vertical position of is fixed to the observations prediction and residual respectively. The current projection basis is depicted on the bottom, where the width of the bar is mapped to variables contribution to the horizontal axis. The starting frame is the normalized SHAP values of the primary observation. In figure \@ref(fig:cheemtourclass) we illustrate the classification case with simulated data; 3 clusters separated in the first 2 dimensions with 2 dimensions of noise.

@@

```{r casepenguins, echo=F, out.width = "100%", fig.cap = "Species classification of Palmer penguin data."}
knitr::include_graphics("./figures/case_penguins.png")
```

@@

```{r casechocolates, echo=F, out.width = "100%", fig.cap = "Chocolates data  type classification (milk or dark). "}
knitr::include_graphics("./figures/case_chocolates.png")
```

@@

```{r casefifa, echo=F, out.width = "100%", fig.cap = "FIFA 2020, regressing log wages [2020 Euros] from aggregations of skill measurements. The primary observation is star offensive player (L. Messi) as compared with a top defensive player (V. van Dijk)."}
knitr::include_graphics("./figures/case_fifa.png")
```

@@

```{r caseames, echo=F, out.width = "100%", fig.cap = "Ames housing 2018 regressing log sales price [2018 USD]."}
knitr::include_graphics("./figures/case_ames2018.png")
```
